Mortality Trend Among Shelter Residents: 

### 2025-06-20 paper/paper.qmd
changed paper to python

```{r}
#| include: false
#| warning: false
#| message: false

# load necessary package
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(gridExtra)

# load data
Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))
```

@@

\newpage
```{r}
#| label: tbl-Table1
#| tbl-cap: Variable In The Data - Detah Among Shelter Residents (2007-2024)
#| echo: false

# Remove the 'id' column
AvgAge_data_no_id <- AvgAge_data %>%
  select(-id)

# Display the table without the 'id' column
kable(AvgAge_data_no_id,
  col.names = c(
    "Year", "Total Decedents",
    "Avg Age of Death (All)", "Male Decedents",
    "Avg Age of Death (Male)", "Female Decedents",
    "Avg Age of Death (Female)", "Trans/NB/2S Decedents",
    "Avg Age of Death (Trans/NB/2S)"
  )
)
```

@@

```{r}
#| label: fig-TotalNumberofDeathByYear
#| fig-cap: Yearly Mortality rate from 2007-2024
#| echo: false

# creating a line graph
AvgAge_data |>
  ggplot(aes(x = year, y = total_decedents)) +
  geom_point() +
  geom_line() +
  labs(title = "Total Shelter Residents Decedents from 2007 to 2024", x = "Year", y = "Total Decedents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = unique(AvgAge_data$year)) +
  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))
```

\newpage
@fig-TotalDecedentbyMonthcovid provides a close-up view of the years 2019 to 2024, covering the period of the COVID-19 pandemic and a bit of the post-pandemic phase. During this time, the number of deaths does not exhibit a clear trend as time progresses. However, in many months during these years, the number of deaths remains notably high, often fluctuating between 10 and 15 deaths per month. Detailed discussion found in @sec-second-dpoint.
```{r}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality rate during COVID (2019-2023)
#| echo: false

# Filter the data for years 2019 to 2022
Death_data_filtered <- Death_data |>
  filter(year >= 2019 & year <= 2023)

# Combine Year and Month into a single column with last two digits of the year
Death_data_filtered$YearMonth <- paste(substr(Death_data_filtered$year, 3, 4), Death_data_filtered$month, sep = "-")

# Convert YearMonth to a factor with ordered levels
Death_data_filtered$YearMonth <- factor(Death_data_filtered$YearMonth, levels = unique(Death_data_filtered$YearMonth))

# Plot the graph for filtered data
Death_data_filtered |>
  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
  geom_line(color = "black") +
  geom_point(color = "black") +
  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) +
  labs(
    title = "Trends in Total Decedents by Month(2019-2023)",
    x = "Month-Year",
    y = "Total Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) +
  theme(plot.title = element_text(hjust = 0.5))
```
\newpage
@fig-T illustrates the number of deaths among genders. The bar graph clearly shows that the number of male decedents is significantly greater than that of female decedents, a trend consistent across all the years presented. Additionally, the graph indicates that the mortality rate among males is more volatile over the years, particularly from 2019 to 2022. These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
```{r}
#| label: fig-T
#| fig-cap: Total Decedents Split by Gender (Female VS Male)
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting
plot_data <- AvgAge_data |>
  select(year, Male_decedents, Female_decedents) |>
  pivot_longer(
    cols = c(Male_decedents, Female_decedents),
    names_to = "Gender",
    values_to = "Mortality"
  )

# Create a new variable for the last two digits of the year with leading zero
plot_data <- plot_data |>
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the bar graph with adjusted year labels
plot_data |>
  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(
    title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
    x = "Year (2000s)",
    y = "Number of Decedents"
  ) +
  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
@fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males. It is important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size. Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph. More information is available in @sec-3-dpoint.

```{r}
#| label: fig-AvgAgeofDecedentsFemaleVSMale
#| fig-cap: Average Age of Death (Female VS Male)
#| echo: false
#| warning: false
#| message: false

# Reshape the data for plotting average age of death
avg_age_data <- AvgAge_data |>
  select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
  pivot_longer(
    cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
    names_to = "Gender",
    values_to = "Average_Age"
  ) |>
  mutate(year_short = sprintf("%02d", year %% 100))

# Create the line graph for average age of death
avg_age_data |>
  ggplot(aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average Age of Death for Male Vs Female Decedents",
    x = "Year",
    y = "Average Age of Death"
  ) +
  scale_color_manual(values = c(
    "Average_age_of_death_male_decedents" = "blue",
    "Average_age_of_death_female_decedents" = "pink"
  )) +
  theme_minimal() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

@@

```{r}
#| label: fig-2007-2010
#| fig-cap: Monthly Mortality rate from 2007-2010
#| echo: false
#| warning: false
#| message: false

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

# Filter data for the years 2007 and 2008
current_year_data_2007 <- Death_data %>% filter(year == 2007)
current_year_data_2008 <- Death_data %>% filter(year == 2008)
current_year_data_2009 <- Death_data %>% filter(year == 2009)
current_year_data_2010 <- Death_data %>% filter(year == 2010)

# Create the bar plots for 2007 and 2008
plot_2007 <- ggplot(current_year_data_2007, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2007",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2008 <- ggplot(current_year_data_2008, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2008",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2009 <- ggplot(current_year_data_2009, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2009",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2010 <- ggplot(current_year_data_2010, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2010",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Arrange plots side by side
grid.arrange(plot_2007, plot_2008, plot_2009, plot_2010, ncol = 2)
```

```{r}
#| label: fig-2011-2014
#| fig-cap: Monthly Mortality rate from 2011-2014
#| echo: false
#| warning: false
#| message: false

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

# Filter data for the years
current_year_data_2011 <- Death_data %>% filter(year == 2011)
current_year_data_2012 <- Death_data %>% filter(year == 2012)
current_year_data_2013 <- Death_data %>% filter(year == 2013)
current_year_data_2014 <- Death_data %>% filter(year == 2014)

# Create the bar plots
plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2011",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2012 <- ggplot(current_year_data_2012, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2012",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2013 <- ggplot(current_year_data_2013, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2013",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2014 <- ggplot(current_year_data_2014, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2014",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Arrange plots side by side
grid.arrange(plot_2011, plot_2012, plot_2013, plot_2014, ncol = 2)
```

```{r}
#| label: fig-2015-2018
#| fig-cap: Monthly Mortality rate from 2015-2018
#| echo: false
#| warning: false
#| message: false

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

# Filter data for the years

current_year_data_2015 <- Death_data %>% filter(year == 2015)
current_year_data_2016 <- Death_data %>% filter(year == 2016)
current_year_data_2017 <- Death_data %>% filter(year == 2017)
current_year_data_2018 <- Death_data %>% filter(year == 2018)

# Create the bar plots

plot_2015 <- ggplot(current_year_data_2015, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2015",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2016 <- ggplot(current_year_data_2016, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2016",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2017 <- ggplot(current_year_data_2017, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2017",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2018 <- ggplot(current_year_data_2018, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2018",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Arrange plots side by side
grid.arrange(plot_2015, plot_2016, plot_2017, plot_2018, ncol = 2)
```

```{r}
#| label: fig-2019-2022
#| fig-cap: Monthly Mortality rate from 2019-2022
#| echo: false
#| warning: false
#| message: false

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

# Filter data for the years

current_year_data_2019 <- Death_data %>% filter(year == 2019)
current_year_data_2020 <- Death_data %>% filter(year == 2020)
current_year_data_2021 <- Death_data %>% filter(year == 2021)
current_year_data_2022 <- Death_data %>% filter(year == 2022)

# Create the bar plots
plot_2019 <- ggplot(current_year_data_2019, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2019",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2020 <- ggplot(current_year_data_2020, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2020",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2021 <- ggplot(current_year_data_2021, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2021",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2022 <- ggplot(current_year_data_2022, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2022",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Arrange plots side by side
grid.arrange(plot_2019, plot_2020, plot_2021, plot_2022, ncol = 2)
```

```{r}
#| label: fig-2023-2024
#| fig-cap: Monthly Mortality rate from 2023-2024
#| echo: false
#| warning: false
#| message: false

# make sure that the month is in order from Jan to Dec.
Death_data$month <- factor(Death_data$month, levels = month.abb)

# Filter data for the years
current_year_data_2023 <- Death_data %>% filter(year == 2023)
current_year_data_2024 <- Death_data %>% filter(year == 2024)

# Create the bar plots
plot_2023 <- ggplot(current_year_data_2023, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2023",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

plot_2024 <- ggplot(current_year_data_2024, aes(x = month, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "black") +
  labs(
    title = "Total Decedents by Month in 2024",
    x = "Month",
    y = "# Decedents"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45)) +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

# Arrange plots side by side
grid.arrange(plot_2023, plot_2024, ncol = 2)
```

>>>>

```{python}
#| include: false
#| warning: false
#| message: false

# load necessary package
import pandas as pd
import matplotlib.pyplot as plt

# load data
Death_data = pd.read_csv("../data/analysis_data/Num_death_analysis_data.csv")
AvgAge_data = pd.read_csv("../data/analysis_data/AvgAge_death_analysis_data.csv")
```

@@

\newpage
```{python}
#| label: tbl-Table1
#| tbl-cap: Variable In The Data - Detah Among Shelter Residents (2007-2024)
#| echo: false

# Remove the 'id' column
AvgAge_data_no_id = AvgAge_data.drop(columns=["id"])

# Rename columns for display
AvgAge_data_no_id.columns = [
    "Year", "Total Decedents",
    "Avg Age of Death (All)", "Male Decedents",
    "Avg Age of Death (Male)", "Female Decedents",
    "Avg Age of Death (Female)", "Trans/NB/2S Decedents",
    "Avg Age of Death (Trans/NB/2S)"
]

# Create the figure and axis
fig, ax = plt.subplots(figsize=(12, len(AvgAge_data_no_id) * 0.4))  # Adjust height to fit rows

# Hide axis
ax.axis("off")

# Create table
table = ax.table(
    cellText=AvgAge_data_no_id.values,
    colLabels=AvgAge_data_no_id.columns,
    loc="center",
    cellLoc="center"
)

# Style the table
table.auto_set_font_size(False)
table.set_fontsize(8)
table.scale(1, 1.5)

# Optional title
plt.title("Table: Variable In The Data – Death Among Shelter Residents (2007–2024)", fontsize=10, pad=20)

plt.tight_layout()
plt.show()

```

@@

```{python}
#| label: fig-TotalNumberofDeathByYear
#| fig-cap: Yearly Mortality rate from 2007-2024
#| echo: false

import matplotlib.pyplot as plt

# Sort by year to ensure correct line order
AvgAge_data_sorted = AvgAge_data.sort_values("year")

# Create the figure and axis
plt.figure(figsize=(8, 5))
plt.plot(AvgAge_data_sorted["year"], AvgAge_data_sorted["total_decedents"], marker="o")

# Set title and labels
plt.title("Total Shelter Residents Decedents from 2007 to 2024", ha="center")
plt.xlabel("Year")
plt.ylabel("Total Decedents")

# Rotate x-axis labels
plt.xticks(AvgAge_data_sorted["year"], rotation=45)

# Set y-axis ticks at intervals of 20
plt.yticks(range(0, AvgAge_data_sorted["total_decedents"].max() + 20, 20))

# Apply minimal style (closest match to theme_minimal in ggplot2)
plt.grid(True, linestyle='--', alpha=0.5)
plt.tight_layout()

# Show plot (Quarto will capture it)
plt.show()

```

\newpage
@fig-TotalDecedentbyMonthcovid provides a close-up view of the years 2019 to 2024, covering the period of the COVID-19 pandemic and a bit of the post-pandemic phase. During this time, the number of deaths does not exhibit a clear trend as time progresses. However, in many months during these years, the number of deaths remains notably high, often fluctuating between 10 and 15 deaths per month. Detailed discussion found in @sec-second-dpoint.
```{python}
#| label: fig-TotalDecedentbyMonthcovid
#| fig-cap: Yearly Mortality Rate During COVID (2019–2023)
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import pandas as pd

# Filter the data for years 2019 to 2023
Death_data_filtered = Death_data[(Death_data["year"] >= 2019) & (Death_data["year"] <= 2023)].copy()

# Ensure month is ordered and standardized
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data_filtered["month"] = pd.Categorical(Death_data_filtered["month"], categories=month_order, ordered=True)

# Sort by year and month
Death_data_filtered = Death_data_filtered.sort_values(by=["year", "month"])

# Combine Year and Month into a "YY-MMM" format
Death_data_filtered["YearMonth"] = Death_data_filtered["year"].astype(str).str[-2:] + "-" + Death_data_filtered["month"].astype(str)

# Plot the data
plt.figure(figsize=(12, 5))
plt.plot(Death_data_filtered["YearMonth"], Death_data_filtered["total_decedents"], marker="o", color="black")

# Show x-axis labels every 6 months
xtick_locs = range(0, len(Death_data_filtered), 6)
xtick_labels = Death_data_filtered["YearMonth"].iloc[xtick_locs]
plt.xticks(ticks=xtick_locs, labels=xtick_labels, rotation=90, fontsize=8)

# Labels and title
plt.title("Yearly Mortality Rate During COVID (2019–2023)", ha="center")
plt.xlabel("Month-Year")
plt.ylabel("Total Decedents")

# Minimal grid styling
plt.grid(True, linestyle="--", alpha=0.5)
plt.tight_layout()

plt.show()

```
\newpage
@fig-T illustrates the number of deaths among genders. The bar graph clearly shows that the number of male decedents is significantly greater than that of female decedents, a trend consistent across all the years presented. Additionally, the graph indicates that the mortality rate among males is more volatile over the years, particularly from 2019 to 2022. These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
```{python}
#| label: fig-T
#| fig-cap: Total Decedents Split by Gender (Female VS Male)
#| echo: false
#| warning: false
#| message: false

import pandas as pd
import matplotlib.pyplot as plt
import numpy as np

# Reshape the data for plotting (from wide to long format)
plot_data = AvgAge_data[["year", "Male_decedents", "Female_decedents"]].copy()
plot_data = plot_data.melt(id_vars="year", value_vars=["Male_decedents", "Female_decedents"],
                           var_name="Gender", value_name="Mortality")

# Create a new variable for the last two digits of the year, with leading zero
plot_data["year_short"] = plot_data["year"] % 100
plot_data["year_short"] = plot_data["year_short"].apply(lambda x: f"{x:02d}")

# Create a grouped bar chart
years = plot_data["year_short"].unique()
x = np.arange(len(years))  # the label locations

male_values = plot_data[plot_data["Gender"] == "Male_decedents"]["Mortality"].values
female_values = plot_data[plot_data["Gender"] == "Female_decedents"]["Mortality"].values

width = 0.35  # width of the bars

fig, ax = plt.subplots(figsize=(10, 5))
ax.bar(x - width/2, male_values, width, label="Male", color="blue")
ax.bar(x + width/2, female_values, width, label="Female", color="pink")

# Labels and formatting
ax.set_xlabel("Year (2000s)")
ax.set_ylabel("Number of Decedents")
ax.set_title("Mortality Numbers of Male Vs Female Decedents over 2007 to 2024", pad=15)
ax.set_xticks(x)
ax.set_xticklabels(years, rotation=0)
ax.legend(title="", loc="lower center", bbox_to_anchor=(0.5, -0.2), ncol=2)
ax.grid(True, axis='y', linestyle='--', alpha=0.5)

plt.tight_layout()
plt.show()

```
@fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males. It is important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size. Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph. More information is available in @sec-3-dpoint.

```{python}
#| label: fig-AvgAgeofDecedentsFemaleVSMale
#| fig-cap: Average Age of Death (Female VS Male) 2007–2024
#| echo: false
#| warning: false
#| message: false

import pandas as pd
import matplotlib.pyplot as plt

# Reshape the data from wide to long format
avg_age_data = AvgAge_data[["year", "Average_age_of_death_male_decedents", "Average_age_of_death_female_decedents"]].copy()
avg_age_data = avg_age_data.melt(
    id_vars="year",
    value_vars=["Average_age_of_death_male_decedents", "Average_age_of_death_female_decedents"],
    var_name="Gender",
    value_name="Average_Age"
)

# Create new column with last two digits of the year
avg_age_data["year_short"] = avg_age_data["year"] % 100
avg_age_data["year_short"] = avg_age_data["year_short"].apply(lambda x: f"{x:02d}")

# Pivot back to format lines cleanly
pivoted = avg_age_data.pivot(index="year_short", columns="Gender", values="Average_Age")

# Plot
plt.figure(figsize=(10, 5))
plt.plot(pivoted.index, pivoted["Average_age_of_death_male_decedents"], marker="o", label="Male", color="blue")
plt.plot(pivoted.index, pivoted["Average_age_of_death_female_decedents"], marker="o", label="Female", color="pink")

# Titles and labels
plt.title("Average Age of Death for Male Vs Female Decedents", ha="center")
plt.xlabel("Year (in 2000s)")
plt.ylabel("Average Age of Death")

# Formatting
plt.xticks(rotation=0)
plt.legend(title="", loc="lower center", bbox_to_anchor=(0.5, -0.25), ncol=2)
plt.grid(True, linestyle="--", alpha=0.5)
plt.tight_layout()

plt.show()
```

@@

```{python}
#| label: fig-2007-2010
#| fig-cap: Monthly Mortality Rate from 2007–2010
#| echo: false
#| warning: false
#| message: false

import pandas as pd
import matplotlib.pyplot as plt

# Ensure month column is a categorical with correct month order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)

# Filter for each year
data_2007 = Death_data[Death_data["year"] == 2007]
data_2008 = Death_data[Death_data["year"] == 2008]
data_2009 = Death_data[Death_data["year"] == 2009]
data_2010 = Death_data[Death_data["year"] == 2010]

# Set up subplot layout
fig, axs = plt.subplots(2, 2, figsize=(12, 8))
axs = axs.flatten()  # Flatten to 1D for easy indexing

# Year-wise loop for consistent plotting
for i, (data, year) in enumerate(zip([data_2007, data_2008, data_2009, data_2010], [2007, 2008, 2009, 2010])):
    axs[i].bar(data["month"], data["total_decedents"], color="black")
    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
    axs[i].set_xlabel("Month")
    axs[i].set_ylabel("# Decedents")
    axs[i].tick_params(axis="x", rotation=45)
    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)

# Layout adjustment
plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-2011-2014
#| fig-cap: Monthly Mortality Rate from 2011–2014
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import pandas as pd

# Ensure months are in correct order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)

# Filter data by year
data_2011 = Death_data[Death_data["year"] == 2011]
data_2012 = Death_data[Death_data["year"] == 2012]
data_2013 = Death_data[Death_data["year"] == 2013]
data_2014 = Death_data[Death_data["year"] == 2014]

# Set up 2x2 subplot layout
fig, axs = plt.subplots(2, 2, figsize=(12, 8))
axs = axs.flatten()

# Plot each year
for i, (data, year) in enumerate(zip([data_2011, data_2012, data_2013, data_2014], [2011, 2012, 2013, 2014])):
    axs[i].bar(data["month"], data["total_decedents"], color="black")
    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
    axs[i].set_xlabel("Month")
    axs[i].set_ylabel("# Decedents")
    axs[i].tick_params(axis="x", rotation=45)
    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)

# Adjust layout for better spacing
plt.tight_layout()
plt.show()
```

```{python}
#| label: fig-2015-2018
#| fig-cap: Monthly Mortality Rate from 2015–2018
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import pandas as pd

# Ensure months are in correct order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)

# Filter data by year
data_2015 = Death_data[Death_data["year"] == 2015]
data_2016 = Death_data[Death_data["year"] == 2016]
data_2017 = Death_data[Death_data["year"] == 2017]
data_2018 = Death_data[Death_data["year"] == 2018]

# Set up 2x2 subplot layout
fig, axs = plt.subplots(2, 2, figsize=(12, 8))
axs = axs.flatten()

# Plot each year's data
for i, (data, year) in enumerate(zip([data_2015, data_2016, data_2017, data_2018], [2015, 2016, 2017, 2018])):
    axs[i].bar(data["month"], data["total_decedents"], color="black")
    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
    axs[i].set_xlabel("Month")
    axs[i].set_ylabel("# Decedents")
    axs[i].tick_params(axis="x", rotation=45)
    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)

# Adjust layout
plt.tight_layout()
plt.show()

```

```{python}
#| label: fig-2019-2022
#| fig-cap: Monthly Mortality Rate from 2019–2022
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import pandas as pd

# Ensure months are in correct order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)

# Filter data by year
data_2019 = Death_data[Death_data["year"] == 2019]
data_2020 = Death_data[Death_data["year"] == 2020]
data_2021 = Death_data[Death_data["year"] == 2021]
data_2022 = Death_data[Death_data["year"] == 2022]

# Set up 2x2 subplot layout
fig, axs = plt.subplots(2, 2, figsize=(12, 8))
axs = axs.flatten()

# Plot each year
for i, (data, year) in enumerate(zip([data_2019, data_2020, data_2021, data_2022], [2019, 2020, 2021, 2022])):
    axs[i].bar(data["month"], data["total_decedents"], color="black")
    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
    axs[i].set_xlabel("Month")
    axs[i].set_ylabel("# Decedents")
    axs[i].tick_params(axis="x", rotation=45)
    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)

# Adjust layout
plt.tight_layout()
plt.show()

```

```{python}
#| label: fig-2023-2024
#| fig-cap: Monthly Mortality Rate from 2023–2024
#| echo: false
#| warning: false
#| message: false

import matplotlib.pyplot as plt
import pandas as pd

# Ensure months are ordered from Jan to Dec
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)

# Filter data for 2023 and 2024
data_2023 = Death_data[Death_data["year"] == 2023]
data_2024 = Death_data[Death_data["year"] == 2024]

# Set up 1x2 subplot layout
fig, axs = plt.subplots(1, 2, figsize=(12, 4))

# Plot each year
for ax, data, year in zip(axs, [data_2023, data_2024], [2023, 2024]):
    ax.bar(data["month"], data["total_decedents"], color="black")
    ax.set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
    ax.set_xlabel("Month")
    ax.set_ylabel("# Decedents")
    ax.tick_params(axis="x", rotation=45)
    ax.grid(True, axis="y", linestyle="--", alpha=0.4)

# Adjust layout
plt.tight_layout()
plt.show()

```
